---
title: "Chapter 9 MCMC"
author: "Stephen R. Proulx"
date: "5/6/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

library(rethinking)
```

Load and process the rugged data
```{r load_process_data}
data(rugged)
d <- as_tibble(rugged)


# make log version of outcome and  extract countries with GDP data
d <- d %>% 
  mutate(log_gdp = log(rgdppc_2000))%>%
  drop_na(rgdppc_2000) %>% 
  mutate(log_gdp_std=log_gdp/mean(log_gdp), rugged_std = rugged/max(rugged) , cid= (cont_africa==1)*1 + (cont_africa==0)*2) %>%
  select(log_gdp_std,rugged_std,cid)
```



```{r}
## R code 9.14
m9.1 <- ulam(
    alist(
        log_gdp_std ~ dnorm( mu , sigma ) ,
        mu <- a[cid] + b[cid]*( rugged_std - 0.215 ) , #link function
        a[cid] ~ dnorm( 1 , 0.1 ) ,
        b[cid] ~ dnorm( 0 , 0.3 ) ,
        sigma ~ dexp( 1 )
    ) , data=d , chains=4 )


show( m9.1 )

## R code 9.15
precis( m9.1 , depth=2 )
```

check the samples in more detail
```{r}


## R code 9.19
pairs( m9.1 )

## R code 9.20
traceplot( m9.1 )

## R code 9.21
trankplot( m9.1 )


```


Run a model with very little data and a very flat prior
```{r}
## R code 9.22
d <- tibble(y=c(-1,1))

set.seed(11)
m9.2 <- ulam(
    alist(
        y ~ dnorm( mu , sigma ) ,
        mu <- alpha ,
        alpha ~ dnorm( 0 , 1000 ) ,
        sigma ~ dexp( 0.0001 )
    ) , data=d , chains=8 , iter=5000)

## R code 9.23
precis( m9.2 )
```
Let's look more closely at what is going on here.
```{r}
pairs(m9.2@stanfit , pars=c("alpha","sigma", "lp__"))
```
```{r}


traceplot( m9.2 )

trankplot( m9.2 )

```


```{r}
## R code 9.24
set.seed(11)
m9.3 <- ulam(
    alist(
        y ~ dnorm( mu , sigma ) ,
        mu <- alpha ,
        alpha ~ dnorm( 1 , 10 ) ,
        sigma ~ dexp( 1 )
    ) , data=d, chains=4 )
precis( m9.3 )


```

Let's look more closely at what is going on here.
```{r}
pairs(m9.3@stanfit , pars=c("alpha","sigma"))
```

```{r}


traceplot( m9.3 )

trankplot( m9.3 )

```


A model that has non-identifiable parameters

```{r}
## R code 9.25
set.seed(41)
y <- rnorm( 100 , mean=0 , sd=1 )

## R code 9.26
set.seed(384)
m9.4 <- ulam(
    alist(
        y ~ dnorm( mu , sigma ) ,
        mu <- a1 + a2 ,
        a1 ~ dnorm( 0 , 1000 ),
        a2 ~ dnorm( 0 , 1000 ),
        sigma ~ dexp( 1 )
    ) , data=list(y=y) , chains=8,iter=2000 )
precis( m9.4 )
```
```{r}
pairs(m9.4@stanfit, pars=c("a1","a2","sigma"))

traceplot( m9.4 )

trankplot( m9.4 )
```

Even though this model has non-identifiable parameters, it gets the big picture correct.  
```{r}
post=as.data.frame(m9.4@stanfit) %>% mutate(asum=a1+a2) 
mcmcpairs(select(post,asum,sigma) , pars=c("asum","sigma"))
```


Now run the same model but with more restrictive priors
```{r}
## R code 9.27
m9.5 <- ulam(
    alist(
        y ~ dnorm( mu , sigma ) ,
        mu <- a1 + a2 ,
        a1 ~ dnorm( 0 , 10 ),
        a2 ~ dnorm( 0 , 10 ),
        sigma ~ dexp( 1 )
    ) , data=list(y=y) , chains=3 )
precis( m9.5 )

```

```{r}
pairs(m9.5@stanfit, pars=c("a1","a2","sigma"))

traceplot( m9.5 )

trankplot( m9.5 )
```

Like before this still gets the total right.
```{r}
post=as.data.frame(m9.5@stanfit) %>% mutate(asum=a1+a2) 
mcmcpairs(select(post,asum,sigma) )

bayesplot::mcmc_areas(select(post,asum,sigma), area_method ="equal height",
            prob=0.975)
```